---
title: How to spell HODL?
author: David Salazar
date: '2018-07-17'
slug: how-spell-hodl
categories:
  - R
tags:
  - finance
  - ggplot2
  - tidyverse
---

# The moody Mr. Market

Anybody who has even a dime on the stock market will eventually get dragged on following the daily (or even hourly) moves in the market. However, this is not only a stressful idea, but also a very ineffective one. Most days in the market won't even bulge your final total return. In fact, final market return is mostly determined by a handful of days alone.

**Let's prove this by comparing the annualized return one would get from investing the last 20 years and being invested 100% of the time, and missing the best days of the market due to bad timing.**

Nota bene: Here I will use price returns from Yahoo!, thus the effect of dividends will not be seen. 

# R!

Let's try to do so in `R` using `tidy` tools all the way: `tidyverse` and `tidyquant`. From downloading the relevant financial data using `tidyquant`, to plotting and more using the packages in the `tidyverse`. 

## Load Packages

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyquant)
```

## Get Data

We will be working with the S&P500:

```{r message=FALSE, warning=FALSE}
sp_500 <- tq_get("^GSPC", get = "stock.prices", from = "1996-12-31", to = "2017-12-29")

total_returns <- sp_500 %>% 
  tq_transmute(select = close, mutate_fun = periodReturn, col_rename = 'total_returns',
               period = "daily") 

lines <- total_returns %>% 
  summarise(mean = mean(total_returns),
            median = median(total_returns),
            perc_99 = quantile(total_returns, probs = 0.99)) %>% 
  gather(key = "statistic", value = "total_returns")

g1 <- total_returns %>% 
  ggplot(aes(x = total_returns)) +
    geom_histogram(fill = "dodgerblue3", binwidth = 0.01, color = "black", alpha = 0.8) +
    geom_vline(data = lines, aes(xintercept = total_returns, color = statistic), linetype = 4) +
    hrbrthemes::theme_ipsum_rc() +
    labs(x = "Total Return for the S&P",
         title = "Daily Returns Histogram") +
    ggthemes::scale_color_gdocs()
g1
```

Even in the histogram we can note that there are a couple of observations with a dominating effect on the overall returns, at both sides of the distribution. 

## Custom Functions

We need a function that, for a given number of best/worst days, returns the final return (both absolute and annualized) of the S&P without using those days in the calculation. 

```{r}
total_absolute_return <- function(tibb, number_days ,initial_investment = 10000) {
  
  added_1 <- tibble((1 + tibb$total_returns)) %>% 
  `colnames<-`("total_returns")

  added_1_date <- timetk::tk_tbl(tibb$date) %>% `colnames<-`("date")

  abs_returns <- cbind(added_1_date, added_1) %>% 
    timetk::tk_tbl() %>% 
    mutate(accumulated = cumprod(total_returns) * initial_investment) %>% 
    tail(1) %>% 
    select(accumulated) %>% 
    mutate(days_taken = number_days)
  
  abs_returns
}

take_days_out <- function(total_returns, number_days, initial = 1, best = TRUE) {
  
  # Define days to take from time series
  if (best == "Both") {
    
    anti_days_1 <- total_returns %>% 
    arrange(desc(total_returns)) %>% 
    slice(initial:number_days)
    
    anti_days_2 <- total_returns %>% 
    arrange(total_returns) %>% 
    slice(initial:number_days)
    
    anti_days <- bind_rows(anti_days_1, anti_days_2)
  } else if (!best) {
    
    anti_days <- total_returns %>% 
    arrange(total_returns) %>% 
    slice(initial:number_days)
    
  } else if (best) {
    
    anti_days <- total_returns %>% 
    arrange(desc(total_returns)) %>% 
    slice(initial:number_days)
  }
  
  # Take those days out
  new_returns <- total_returns %>% anti_join(anti_days, by = "date")
  
  # Calculate annualized return without those days
  ret <- new_returns %>% 
    tq_transmute(select = total_returns, Return.annualized) %>% 
    as_tibble() %>% 
    mutate(days_taken = number_days)
  
  abs_returns <- total_absolute_return(new_returns, number_days = number_days)
  
  ret <- ret %>% left_join(abs_returns, by = "days_taken")
  
  ret 
}
```

```{r}
missing_best_n_days <- function(n, best = TRUE) {
  if (best) {    
    ret <- glue::glue("Missing {n} best days")
  } else {
    ret <- glue::glue("Missing {n} worst days")
  }
  ret
}

missing_action_days <- function(n) {
  glue::glue("Missing {n*2} action days.")
}
```

## Check the results: Best Days

Let's see our results of being so out of luck that you are out of the market precisely the days the market has seen the best results overall:

```{r message=FALSE, warning=FALSE}
c(10, 20, 30) %>% 
  map_df(~ take_days_out(total_returns, number_days = .x)) %>% 
  rbind(take_days_out(total_returns, number_days = 0, 0)) %>% 
  arrange(days_taken) %>% 
  mutate(days_taken = missing_best_n_days(days_taken)) -> results

results %>% 
  rename(market_return = total_returns) %>% 
  ggplot(aes(x = factor(days_taken), y = accumulated)) +
    geom_col() +
    geom_label(aes(label = scales::percent(market_return))) +
    coord_flip() +
  hrbrthemes::theme_ipsum_rc(grid = "Y") +
  labs(x = "",
       y = "Absolute return", 
       subtitle = "A few days determine your return. Annualized Return in percent",
       title = "Price Return S&P 500 for 20 years",
       caption = "Performance of a $10.000 investment between January 1, 1997 and December 31, 2017. Price Return") +
  scale_y_continuous(labels = scales::dollar)
```

You could have followed the market for the whole 20 years, and yet just 20 of those days would have been truly important. Had you missed those 20 days, you would have lost almost everything the market has to offer. **The lesson is quite clear: most days are inconsequential for your returns, and thus paying attention to all of them is quite simply masochistic.**

## Check the results: Worst Days

Let's repeat the exercise, but this time assuming you are going to have the luck of being out of the market at precisely the worst days possible. 

```{r message=FALSE, warning=FALSE}
c(10, 20, 30) %>% 
  map_df(~ take_days_out(total_returns, number_days = .x, best = FALSE)) %>% 
  rbind(take_days_out(total_returns, number_days = 0, 0)) %>% 
  arrange(days_taken) %>% 
  mutate(days_taken = missing_best_n_days(days_taken, best = FALSE)) -> results

results %>% 
  rename(market_return = total_returns) %>% 
  ggplot(aes(x = factor(days_taken), y = accumulated)) +
    geom_col() +
  coord_flip() +
    geom_label(aes(label = scales::percent(market_return))) +
  hrbrthemes::theme_ipsum_rc(grid = "Y") +
  labs(x = "",
       y = "Absolute return", 
       subtitle = "A few days determine your return. Annualized Return in percent",
       title = "Price Return S&P 500 for 20 years",
       caption = "Performance of a $10.000 investment between January 1, 1997 and December 31, 2017. Price Return") +
  scale_y_continuous(labels = scales::dollar)
```

Here the results are much more dramatic: the overall effect of just a few days is huuugee on your final return. Thus, this only creates more emphasis on the final result: most days are totally inconsequential; your overall result is literally determined by just a handful of days. 

## Check the results: Out of the action for both

Now let's combine the last two exercises: an investor that will be out for both periods of time. That is, miss the 10 best days AND the 10 worst days. 

```{r message=FALSE, warning=FALSE}
c(10, 20, 30) %>% 
  map_df(~ take_days_out(total_returns, number_days = .x, best = "Both")) %>% 
  rbind(take_days_out(total_returns, number_days = 0, 0)) %>% 
  arrange(days_taken) %>% 
  mutate(days_taken = missing_action_days(days_taken)) -> results

results %>% 
  rename(market_return = total_returns) %>% 
  ggplot(aes(x = factor(days_taken), y = accumulated)) +
    geom_col() +
  coord_flip() +
    geom_label(aes(label = scales::percent(market_return))) +
  hrbrthemes::theme_ipsum_rc(grid = "Y") +
  labs(x = "",
       y = "Absolute return", 
       subtitle = "A few days determine your return. Annualized Return in percent",
       title = "Price Return S&P 500 for 20 years",
       caption = "Performance of a $10.000 investment between January 1, 1997 and December 31, 2017. Price Return") +
  scale_y_continuous(labels = scales::dollar)
```

Given the assymetry in returns (the worst worst days are greater in absolute value than the best best days), missing out of the best days in exchange of missing out in the worst days results in an overall improvement on the result. 

## Check the results: Colombian Market

Let's re-do the experiment, but this time for the Colombian Market using all the data available from Yahoo. 

```{r}
icol <- tq_get("ICOL", get = "stock.prices")

total_returns <- icol %>% 
  tq_transmute(select = adjusted, mutate_fun = periodReturn, col_rename = 'total_returns',
               period = "daily") 

lines <- total_returns %>% 
  summarise(mean = mean(total_returns),
            median = median(total_returns),
            perc_99 = quantile(total_returns, probs = 0.99)) %>% 
  gather(key = "statistic", value = "total_returns")

g1 <- total_returns %>% 
  ggplot(aes(x = total_returns)) +
    geom_histogram(fill = "dodgerblue3", binwidth = 0.01, color = "black", alpha = 0.8) +
    geom_vline(data = lines, aes(xintercept = total_returns, color = statistic), linetype = 4) +
    hrbrthemes::theme_ipsum_rc() +
  coord_flip() +
    labs(x = "Total Return for the iCOLCAP. ",
         title = "Daily Returns Histogram") +
    ggthemes::scale_color_gdocs()
g1
```

```{r message=FALSE, warning=FALSE}
c(10, 20, 30) %>% 
  map_df(~ take_days_out(total_returns, number_days = .x)) %>% 
  rbind(take_days_out(total_returns, number_days = 0, 0)) %>% 
  arrange(days_taken) %>% 
  mutate(days_taken = missing_best_n_days(days_taken)) -> results

results %>% 
  rename(market_return = total_returns) %>% 
  ggplot(aes(x = factor(days_taken), y = accumulated)) +
    geom_col() +
    geom_label(aes(label = scales::percent(market_return))) +
  hrbrthemes::theme_ipsum_rc(grid = "Y") +
  labs(x = "",
       y = "Absolute return", 
       subtitle = "A few days determine your return. Annualized Return in percent",
       title = "Price Return iCOLCAP for 5 years",
       caption = "Performance of a $10.000 investment between in the last 5 years. Price Return") +
  scale_y_continuous(labels = scales::dollar)
```

The overall return has been disastrous (thanks Oil prices!) but it could have been much, much worse had you missed just a few days. 

Let's check the worst days:

```{r message=FALSE, warning=FALSE}
c(10, 20, 30) %>% 
  map_df(~ take_days_out(total_returns, number_days = .x, best = FALSE)) %>% 
  rbind(take_days_out(total_returns, number_days = 0, 0)) %>% 
  arrange(days_taken) %>% 
  mutate(days_taken = missing_best_n_days(days_taken, best = FALSE)) -> results

results %>% 
  rename(market_return = total_returns) %>% 
  ggplot(aes(x = factor(days_taken), y = accumulated)) +
    geom_col() +
  coord_flip() +
    geom_label(aes(label = scales::percent(market_return))) +
  hrbrthemes::theme_ipsum_rc(grid = "Y") +
  labs(x = "",
       y = "Absolute return", 
       subtitle = "A few days determine your return. Annualized Return in percent",
       title = "Price Return iCOLCAP for 5 years",
       caption = "Performance of a $10.000 investment between in the last 5 years. Price Return") +
  scale_y_continuous(labels = scales::dollar)
```

Just missing 10 god-forsaken days in the Market would have take us from a -7.0% annualized return to a 5.8%!. Again, the overall market result is determined by just a handful of days. 

# Conclusion

The overall return of your investmen is always going to be determined by just a handful of days. Worrying about the performance of the market every day is not only stressful but also ineffective. 

