---
title: "Netflix Habits through data"
author: "David Salazar"
date: '2019-10-15'
slug: netflix-habits-through-data
tags: []
categories: []
---

# Netflix Habits

In the past, I believe I have spent an inordinate amount of time watching series and movies on Netflix. To try to gauge how my habits have changed through time, I downloaded the data that Netflix makes available and, of course, used ´R´ to analyze it. 

## Tidy Tools

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tsibble)
```


```{r include=FALSE}
library(extrafont)
extrafont::loadfonts()
```


Let's have a first look:

```{r message=FALSE, warning=FALSE, include=FALSE}
data <- read_csv("NetflixViewingHistory.csv") %>% mutate(Date = lubridate::mdy(Date)) %>% as_tsibble(key = Title) 
data
```

Sadly, there's not that much information. However, let's try to gauge how many shows have I watched:

### By month

```{r}
data %>% index_by(year_month = yearmonth(Date)) %>% summarise(n = n()) %>%
  ggplot(aes(x = year_month, y = n)) +
    geom_col(fill = "#e50914") + hrbrthemes::theme_ipsum_rc() + labs(x = "Date", y = "Number of Episodes or Movies watched",
                                                     title = "Netflix usage per Month")
```

Clearly, something was going on through 2014 and 2017. Both were pivotal moments in my life, and both were times were I watched way too many shows :(.

### By year

```{r}
data %>% index_by(year_month = lubridate::year(Date)) %>% summarise(n = n()) %>%
  ggplot(aes(x = year_month, y = n)) +
    geom_col(fill = "#e50914") + hrbrthemes::theme_ipsum_rc() + labs(x = "Date", y = "Number of Episodes or Movies watched",
                                                     title = "Netflix usage per Year")
```

### Most watched series

The way the data is stored is the following ´SeriesName:episode´. Let's see which were the most watched shows over the years using some ´regex´ magic thanks to Google. 

```{r}
data %>% 
  mutate(is_series = stringr::str_detect(Title, ":")) %>% 
  filter(is_series == TRUE) %>% 
  select(-is_series) %>% 
  mutate(series_name = stringr::str_extract(Title, "^[^:]+:\\s*")) %>% 
  index_by(year = lubridate::year(Date)) %>% 
  group_by(series_name) %>% 
  summarise(episodes_watched = n()) %>% 
  index_by(year) %>% 
  arrange(desc(episodes_watched)) %>% 
  slice(1) %>% 
  kableExtra::kable() %>% 
  kableExtra::kable_styling()

```

Apparently, the 2014 and 2017 frenzy were powered by Grey's Anatomy and Friends, two shows that I am not so proud to have watched xD. 


